### 2022-09-07 AISTATS2023/paper.tex
started method description

@@

\section{THE U-DALE METHOD}

\subsection{Uncertainty Quantification}

The uncertainty of the global explanation emerges from the natural characteristics of the experiment, i.e.,~the data generating distribution and the black-box function.

@@

\subsubsection*{References}

@@

>>>>

\newcommand{\dfdx}{\frac{\partial f}{\partial x_s}}
\newcommand{\xc}{\mathbf{x_c}}
\newcommand{\DY}{\mathbf{\Delta Y}}
\newcommand{\xb}{\mathbf{x}}
\newcommand{\Xcb}{\mathcal{X}_c}
\newcommand{\Xb}{\mathcal{X}}

@@

\paragraph{Notation.}

We refer to random variables (rv) using uppercase \( X \), whereas to
simple variables with plain lowercase \( x \). Bold denotes a vector;
\( \xb \) for simple variables or \(\mathbf{X}\) for rvs. Often, we
partition the input vector \(\xb \in \mathbb{R}^D\) to the feature of
interest \(x_s \in \mathbb{R} \) and the rest of the features
\(\xc \in \mathbb{R}^{D-1}\). For convenience we denote it as
\((x_s, \mathbb{x}_c)\), but we clarify that it corresponds to the vector
\((x_1, \cdots , x_s, \cdots, x_D)\). Equivalently, we denote the
corresponding rv as \(X = (X_s, \mathbf{X}_c)\). The black-box
function is \(f : \mathbb{R}^D \rightarrow \mathbb{R}\) and the
feature effect of the \(s\)-th feature is
\(f_{\mathtt{<method>}}(x_s)\), where \(\mathtt{<method>}\) is the
name of the feature effect method. \footnote{An extensive list of all
  symbols used in the paper is provided in the helping material.}

\paragraph{Feature Effect Methods.} PDPlots formulate the feature
effect of the \(s\)-th attribute as an expectation over the marginal
distribution \(\mathbf{X}_c\), i.e.,
\(f_{\mathtt{PDP}}(x_s) =
\mathbb{E}_{\mathbf{X}_c}[f(x_s,\mathbf{X}_c)]\). MPlots formulate
it as an expectation over the conditional
\(\mathcal{X}_c|\mathcal{X}_s\), i.e.,
\(f_{\mathtt{MP}}(x_s) = \mathbb{E}_{\mathcal{X}_c|\mathcal{X}_s = x_s}[f(x_s,
\mathcal{X}_c)]\). ALE computes the global effect at \(x_s\) as an
integration of local effects. The local effects are measured as the
expected change on the output
\( \frac{\partial f(x_s, \mathcal{X}_c)}{\partial x_s} \) over the
conditional distribution \( \Xcb|\mathcal{X}_s\). The formula that
defines ALE is presented below:

\begin{equation}
  \label{eq:ALE}
  f_{\mathtt{ALE}}(x_s) = c + \int_{-\infty}^{x_s} \mathbb{E}_{\Xcb|\mathcal{X}_s=z}\left[\frac{\partial f(z, \mathcal{X}_c)}{\partial z}\right] \partial z
\end{equation}
%
The constant \(c\) is used for centering the ALE plot. PDPlots
integrate over unrealistic instances due to the use of the marginal
distribution \( p(\mathcal{X}_1) \). Therefore, they incorrectly
result in a quadratic effect in the region \(x_1 \in [0, 1]\). MPlots
resolve this issue using the conditional distribution
\( \mathcal{X}_2|\mathcal{X}_1 \) but suffer from computing combined
effects. ALE plots resolve, both they have two drawbacks TODO add more
info.

\paragraph{Qunatify the Heterogeneous Effects.}
It is crucial for feature effect methods to inform about the
heterogeneous effects. Elaborate. Interpretation of the heterogeneous
effects behind the global effect is available only for PDP, with three
different approaches; (a) ICE and d-ICE plots provide a visual
understanding of the heterogeneous effects. (b) grouping of ICE in
homogeneous clusters, for spliting the input space into subspace(s)
with homogeneous effects (c) Feature Interaction strength indexes,
like H-statistic, provide a value indicating how much a feature
interacts with the others (not the type of interaction). There is no
method for quantifying the heterogeneous effects, based on
ALE. Therefore, no method to exploit the advantages of ALE while, on
the same time, informing about the heterogeneous effects.

\paragraph{Bin Spliting.} ALE also has the peculiarity of spliting the
axis into intervals, allocating the instances of the training set in
the intervals and compute a single (constant) effect in each
interval. With DALE extension, bin spliting is decoupled from instant
effect estimation. With our extension for measuring the heterogeneous
effects, we transfrom interval spliting from a step to a clustering
problem with a meaningful objective to minimise. We provide a thorough
analysis, where we show that our objective has a consistent
meaning. It can be split in two parts; the first part is the
unavoidable uncertainty due to the natural characteristics of the
experiment, i.e.,~the data generating distribution and the black-box
function. The second part is an added uncertainty due to the
limited-samples estimation, that enforces to create groups with
constant main effect. We opt for minimizing the objective, i.e. sum of
the two uncertainties, that given that the first uncertainty is
independent of the bin spliting, therefore we want to minimize the
added uncertainty. To conclude, we transform the axis-spliting into an
unsupervised clustering problem with a principled objective. We a
computationally-grounded solution that works out-of-the-box, relaxing
the user from providing a hyperparameter without any indication which
one is the correct. This step can be used independently of whether the
user wants to explore the heterogeneous effects or not.

\section{THE ... METHOD}

\subsection{ALE with Uncertainty Quantification}

ALE defines the local effect of \(x_s\) on \(f(\cdot)\) at
\((x_s, \xc)\) as \(\dfdx (x_s, \xc)\). Given that the black-box
function \(f(\cdot)\) is a deterministic predictor, the local effect
is also a deterministic quantity. Global methods summarize the local
explanations into a single quantity. Therefore, ALE summarizes all the
local explanations at \(x_s\) by averaging the local effects accross
all values of \(\xc\) weighting them by \(p(\xc|x_s)\). In other
words, ALE globalizes the local explanations at \(x_s\) through the
expected change in the output \(y\) wrt \(x_s\):

\begin{equation}
  \label{eq:ALE_mean}
  \mu(x_s) = \mathbb{E}_{\xc|x_s}\left [\dfdx (x_s, \xc)\right ]
\end{equation}
%
As described at the Introduction, limiting the explanation at this
level does not shed light to possible heterogeneous effects behind the
averaged explanation. Therefore, we model the model the uncertainty of
the effect at \(x_s\) as the variance of the local explanations:

\begin{equation}
  \label{eq:ALE_var}
  \sigma^2(x_s) = \mathrm{Var}_{\xc|x_s}\left [\dfdx (x_s, \xc) \right ]
\end{equation}

\noindent
%
The uncertainty of the explanation emerges from the natural
characteristics of the experiment, i.e.,~the data generating
distribution and the black-box function. It is important to clarify
that the variance is only a way to model the uncertainty. Other
statistical properties can also be used. Elaborate.

ALE computes the final effect at \(x_s\) by accumulating/integrating
the averaged local effects \(\mu (x_s=z) \) over all values of \(z\)
from \(x_{min}\) until \(x_s\), as show in Eq.~(\ref{eq:ALE}). The
choice for \(x_{min}\) is not important, as it only affects the
centering of ALE plot along the vertical axis. Equivalently, we define
the accumulated uncertainty until the point \(x_s\), as the integral
of the variance of local effects:

\begin{equation}
  \label{eq:3}
  f_{\mathtt{ALE}}^{\sigma^2}(x_s) = \int_{z_1, min}^{x_s} \sigma^2(z) \partial z
\end{equation}

In ALE plots the absolute value at a specific point is not an
interpretable quantity (add ref to paper). The meaningful
interpretation is the effect at a specific point, i.e. what happens to
the output given a small change in the feature of interest. The
integration takes place only for making the visual interpretation
smooth, for example, to spot larger intervals where the effect of a
\textbf{local} change is continuously negative. The same stands for
the accumulated uncertainty. It is meaningful to measure, only, the
uncertainty at at a specific point to understand the heterogeneous
effects behind the averaged effect. In the experimental Section, we
propose appropriate visualizations to concentrate the user to this
effect. However, the accumulated uncertainty, i.e. the aggregate
uncertainty allong all the points of the plot, will help us define a
meaningfull objective for the interval spliting step. (TODO: check
here if the idea about whether the accumulated uncertainty is a good
metric for modeling the interaction strength)

\subsection{Uncertainty Quantification at an interval}

In real scenarios, we have ignorance about the data-generating
distribution \(p(x_s, \mathbf{x}_c)\) and on estimations based on the
limited instances of the training set. Therefore, it is impossible to
estimate Eqs.~\eqref{eq:ALE_mean},~\eqref{eq:ALE_var} at the
granularity of a point \(x_s\)\, since the possibility to observe a
sample in the interval \([x_s - h, x_s + h]\) is zero, when
\(h \to 0\). Therefore, for estimation purposes, we are obliged to
create larger intervals (\(h>0\)). The mean effect at the interval
\([z_1, z_2)\) is the mean of the expected effects:

\begin{equation}
  \label{eq:mu_bin}
  \mu(z_1, z_2) = \frac{1}{z_2 - z_1} \int_{z_1}^{z_2}
  \mathbb{E}_{\xc|x_s=z}\left [\frac{\partial f}{\partial x_s} \right ] \partial z
\end{equation}

\noindent
Accordingly, the accumulated uncertainty (variance) at the interval
\([z_1, z_2)\) is:

\begin{equation}
  \label{eq:var_bin}
  \sigma^2(z_1, z_2) = \int_{z_1}^{z_2}
  \mathbb{E}_{\xc|x_s=z} \left [ (\frac{\partial
      f}{\partial x_s} - \mu(z_1, z_2) )^2 \right] \partial z
\end{equation}

\noindent

\paragraph{Theorem 1.}

The accumulated variance at the interval \([z_1, z_2)\) is the
accumulation of the all variances inside the interval plus the
accumulated squared residuals:

\begin{equation}
 \sigma^2(z_1, z_2) = \int_{z_1}^{z_2} \sigma^2(z) + \rho^2(z) \partial z
\end{equation}
%
where \(\rho(z) = \mu(z) - \mu(z_1, z_2)\). The proof is at the
Appendix. Theorem 1 decouples the (observable) variance in the
interval \([z_1, z_2)\) into two terms. The first term
\(\int_{z_1}^{z_2} \sigma^2(z) \partial z\) accumulates the variance
of the local explanations (Eq.~(\ref{eq:ALE_var})) for all points
inside the interval. The second term accumulates the squared
differences between the mean effect at the interval and mean of the
local explanations for all points inside the interval. The first term
accumulates the the uncertainty due to the natural characteristics of
the experiment, whereas, the second term adds extra uncertainty due to
limiting the granularity of the effect at the bin level.

\subsection{Estimation of Effect at an Interval}

As stated before, Eqs.~\eqref{eq:mu_bin},~\eqref{eq:var_bin} can be
estimated from the set \(\mathcal{S}\) of the instances of the
training set with the \(s\)-th feature lying inside the interval,
i.e., \( \mathcal{S}= \{ \mathbf{x}^i : z_1 \leq x^i_s < z_2 \}
\). The mean effect at the interval, Eq.~(\ref{eq:mu_bin}), can be approximated
by:

\begin{equation}
  \label{eq:mean_estimation}
  \hat{\mu}(z_1, z_2) = \frac{1}{|\mathcal{S}|} \sum_{i:\mathbf{x}^i \in
    \mathcal{S}} \left [ \dfdx(\mathbf{x}^i) \right ]
\end{equation}

and the accumulated variance at the interval, Eq.~(\ref{eq:var_bin})
can be approximated by

\begin{equation}
  \label{eq:variance_estimation}
  \hat{\sigma}_k(z_1, z_2) = \frac{1}{|\mathcal{S}|} \sum_{i:\mathbf{x}^i \in
  \mathcal{S}} \left ( \dfdx(\mathbf{x}^i) - \hat{\mu}_k(z_1, z_2) \right )^2
\end{equation}

The approximation is unbiased only if the points are uniformly
distributed in \([z_{k-1}, z_k]\). (TODOs: Check what happens
otherwise).

\paragraph{Uncertainty of the global effect.}

Eq.~\eqref{eq:variance_estimation} gives an approximation of the
uncertainty of the bin effect.
The uncertainty of the global effect is
simply the sum of the uncertainties in the bin effects.

\paragraph{Minimizing the uncertainty}

Solving the problem of finding (a) the optimal number of bins \(K\) and (b) the optimal bin limits for each bin \([z_{k-1}, z_k] \forall k\) to minimize:

\begin{equation}
  \label{eq:1}
  \mathcal{L} = \sum_{k=0}^K \hat{\sigma}_k(z_{k-1}, z_k)
\end{equation}
%
The constraints are that all bins must include more than \(\tau\)
points, i.e., \(|\mathcal{S}_k| \geq \tau\).

\noindent
TODOS. Show theoretically that \(\mathcal{L} \geq \int_{x_{s, \min}}^{x_{s, \max}}\sigma^2(x_s) \partial x_s\)

@@

@@

\section*{Appendix}

\subsection{Proof for variance of the bin}
